---
title: "Lab 1 Notes"
author: Jose Toledo Luna
toc: false
date: today
---

## Learning Objectives

- Develop basic R programming skills (e.g., creating vectors, performing operations, installing and using packages) 
 ￼
- Apply fundamental statistical functions such as mean, median, minimum, maximum, and standard deviation  ￼

## Notes

We will use the following notes to cover material for Lab 1

1. [Fundamentals of R](https://ucla-stats13.netlify.app/notes/01-fundamentals)
2. [Stats 10 Lab 1 Handout](https://app.box.com/s/af91nv5gy97atmdt6uwsghnzce6znfmd)

<!-- 2. [Importing Data in R](https://ucla-stats13.netlify.app/notes/02-importing_data)
3. [Summarizing Numerical Data](https://ucla-stats13.netlify.app/notes/03-intro_numerical_variables)
4. [Summarizing Categorical Data](https://ucla-stats13.netlify.app/notes/04-intro_categorical_variables) -->


## Questions

1.  Create a code chunk below this question and print the statement:
    "`My name is (your name) and my UID is (your UID). I am looking forward to Stats 10!`"

2.  Create a vector named `titles` and fill it with at least 3 book
    titles. Print the vector `titles`.

3.  Run the code chunk below using `titles`. Read the error message and
    explain why an error occurs.

```{r, eval=FALSE}
sum(titles)
```

4.  Create a vector named vector_log and make it hold the values `TRUE`,
    `FALSE, TRUE, FALSE`. Try multiplying the values by 2 and using the
    `mean()` function. Explain your results and how R treats logical
    values.


5.  You are given a vector containing the scores of 50 students on a
    recent exam. Find the minimum, maximum, mean, and standard deviation
    of the scores. Print each value (for example, print "Mean: \_\_\_").

```{r}
# You do not have to change this code
set.seed(123456789)
scores <- sample(0:100, 50, replace = TRUE)
print(scores)
```


6.  Using the `scores` vector, find the mean value without using the
    `mean()` function. (Hint: The mean is calculated by finding the sum of all values and dividing by the total number of values.)


-   Explain whether there is a difference between the means found in #5
    and #6 and why it does or does not exist.

7.  Knit the file into a PDF and name it "*yourname_yourUID Stats 10 Lab
    1*". Submit the file into Canvas.




