```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  warning = FALSE,
  eval = TRUE
)
```

The following packages were used for this tutorial but are not required for your lab

<details>

<summary>Show Code</summary>

```{r}
library(ggplot2)
library(patchwork)
```

```{r}
theme_set(theme_bw())
theme_replace(panel.grid.minor = element_blank(),
              panel.grid.major = element_blank())
```

</details>

## Binomial Distribution

The binomial distribution is a discrete probability distribution. It describes the outcome of $n$ independent trials in an experiment. Each trial is assumed to have only two outcomes, either success or failure. If the probability of a successful trial is $p$, then the probability of having $x$ successful outcomes in an experiment of $n$ independent trials is as follows

$$
P(X = x) = \frac{n!}{(n-x)! \, x!} p^x (1-p)^{n-k}
$$
where $x =0,1,2,\dots,n$

Let $X \sim Bin(n,p)$ be a random variable following a binomial distribution with mean $np$ and standard deviation $\sqrt{np(1-p)}$. In R, the following functions described in the table below, allows us to summarize the function relating to the binomial distribution

+------------+------------------------------------------------+
| Function   | Description                                    |
+============+================================================+
| dbinom     | Binomial density\                              |
|            | (Probability Density Function)                 |
+------------+------------------------------------------------+
| pbinom     | Binomial distribution\                         |
|            | (Cumulative Distribution Function)             |
+------------+------------------------------------------------+
| qbinom     | Quantile function of the Binomial distribution |
+------------+------------------------------------------------+
| rbinom     | Binomial random number generation              |
+------------+------------------------------------------------+




## dbinom



## pbinom


## qbinom


## rbinom

