---
toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  warning = FALSE,
  eval = TRUE)
```

## Overview of Confidence Intervals for a Proportion



## Dataset Description

We will use the `pawnee` dataset in our class website

```{r}
pawnee <- read.csv(file="/Users/toledo60/Desktop/Projects/personal-website/ucla/stats10/data/pawnee.csv")
```

A brief description of the dataset


> The Sweetums candy factory in Pawnee, Indiana, is under investigation for violating EPA regulations. Factory workers have improperly disposed of arsenic and sulfur waste from the candy-making process, and the contamination has reached the local water supply! We have data for arsenic and sulfur levels from the water in all houses within a 2-mile radius of the factory


Some important variables include:

* `Arsenic`: arsenic levels for each home in ppm
* `Sulfur`: sulfur levels for each home in ppm
* `New_hlth_issue`: Indicates `Y` is someone living at the home has experienced a major health issue after the date of contamination, `N` if no new health complications









