---
toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  warning = FALSE,
  eval = TRUE)
```

## Overview of Confidence Intervals for a Proportion


A point estimate is a single quantity estimating the population parameter. It would be more ideal to give a range of plausible values for the parameter instead with some given *confidence*.

Recall a *point estimate* for the population proportion is the sample proportion $\hat{p}$:
$$
\hat{p} \rightarrow p
$$
An *interval estimate* for the population proportion will then be of the form:
$$
(LB,UB) \rightarrow p
$$

where the lower bound is 

*LB = point estimate - margin of error* 

Similarly, the upper bound is 

*UB = point estimate + margin of error*

A margin of error consists of the standard error and a critical value determined by the confidence level.
The margin of error can be defined as 

*Margin of Error = Critical Value x Standard Error*

The critical value that we use may be obtained from a $t$-distribution if the assumptions necessary for the Central Limit Theorem *(CLT)* are not valid. Alternatively, if these assumptions hold, we can obtain the critical value from a standard normal distribution.


To verify the assumptions to apply CLT, we must have at least 10 successes and 10 failures
$$
n \hat{p} \geq 10 \quad \& \quad n(1-\hat{p}) \geq 10
$$
Given the assumption hold, the margin of error can be calculated as 
$$
ME = z^* \times SE
$$
where $z^*$ is a value obtained from a standard normal distribution given a certain confidence level.
Typical confidence levels include $90\%,95\%,99\%$, We can visualize how to obtain the critical values $z^*$ for these values as follows

<details>
<summary>Show Code</summary>

This code is for teaching purposes only, you do not need them for your lab

```{r}
library(ggplot2)
library(patchwork)
```


```{r}
theme_set(theme_bw())
theme_replace(panel.grid.minor = element_blank(),
              panel.grid.major = element_blank())
```

```{r}
x_limits <- c(-4,4)

p1 <- ggplot(data.frame(x = x_limits), aes(x)) +
  stat_function(fun = dnorm,
                geom = "line",
                xlim = x_limits) +
  stat_function(fun = dnorm,
                geom = "area",
                fill = 'steelblue',
                alpha =0.3,
                xlim = c(-1.65, 1.65))+
  geom_vline(xintercept = 1.65,linetype=2,color = 'steelblue')+
  geom_vline(xintercept = -1.65,linetype=2,color = 'steelblue')+
  labs(x = '',y= 'Density')+
  annotate("text", x = 0, y = 0.1, 
           label = paste0(90,"%"),
           size=4.5)+
  scale_x_continuous(name = '',limits = x_limits,
                     breaks = -4:4)

p2 <- ggplot(data.frame(x = x_limits), aes(x)) +
  stat_function(fun = dnorm,
                geom = "line",
                xlim = x_limits) +
  stat_function(fun = dnorm,
                geom = "area",
                fill = '#cf5d55',
                alpha =0.3,
                xlim = c(-1.96, 1.96))+
  geom_vline(xintercept = 1.96,linetype=2,color = '#cf5d55')+
  geom_vline(xintercept = -1.96,linetype=2,color = '#cf5d55')+
  labs(x = '',y= 'Density')+
  annotate("text", x = 0, y = 0.1, 
           label = paste0(95,"%"),
           size=4.5)+
  scale_x_continuous(name = '',limits = x_limits,
                     breaks = -4:4)

p3 <- ggplot(data.frame(x = x_limits), aes(x)) +
  stat_function(fun = dnorm,
                geom = "line",
                xlim = x_limits) +
  stat_function(fun = dnorm,
                geom = "area",
                fill = '#71c7b0',
                alpha =0.3,
                xlim = c(-2.58, 2.58))+
  geom_vline(xintercept = -2.58,linetype=2,color = '#71c7b0')+
  geom_vline(xintercept = 2.58,linetype=2,color = '#71c7b0')+
  labs(x = '',y= 'Density')+
  annotate("text", x = 0, y = 0.1, 
           label = paste0(99,"%"),
           size=4.5)+
  scale_x_continuous(name = '',limits = x_limits,
                     breaks = -4:4)

```

</details>


```{r,echo=FALSE,  fig.width = 8,fig.height = 5}
p1 + p2 +p3 + plot_layout(ncol = 3)
```

* For a 90\% CI use $z^* = 1.65$
* For a 95\% CI use $z^* = 1.96$
* For a 99\% CI use $z^* = 2.58$

In general, for a $100 \times (1-\alpha)\%$ CI the correspond critical value will be $z_{1-\alpha/2}^*$ which can then easily be computed using `qnorm`. For example, if for some reason we were interested in a 
$78\%$ CI = $100 \times (1-0.22)$ the corresponding critical value would be $z_{1-0.22/2}$

```{r}
qnorm(1-0.22/2)
```

We can verify this with the usual confidence interval levels:

For a 90\% CI use $z^* = 1.65$
```{r}
qnorm(1-0.1/2)
```

For a 95\% CI use $z^* = 1.96$
```{r}
qnorm(1-0.05/2)
```

For a 99\% CI use $z^* = 2.58$
```{r}
qnorm(1-0.01/2)
```


Now that we know how to obtain the critical value, all that is left to do is compute the standard error. The standard error is 
$$
SE(\hat{p}) = \sqrt{\frac{p(1-p)}{n}}
$$
Therefore, a $100(1-\alpha)\%$ CI becomes
$$
\begin{align*}
100(1-\alpha)\% \, CI &=  
(LB,UB) \\[5pt]
&= (\hat{p} - ME, \hat{p}+ME) \\[5pt]
&= (\hat{p} - z_{1-\alpha/2} \times SE, \, \, \hat{p} + z_{1-\alpha/2} \times SE ) \\[5pt]
&= \left(\hat{p} - z_{1-\alpha/2}^* \, \sqrt{\frac{p(1-p)}{n}}, \,\,
\hat{p} + z_{1-\alpha/2}^* \, \, \sqrt{\frac{p(1-p)}{n}} \right)
\end{align*}
$$

**Interpretation**

> If repeated samples were taken and the $95\%$ confidence interval computed for each sample, $95\%$ of the intervals would contain the population parameter. Naturally, $5\%$ of the intervals would not contain the population parameter


**Procedure**

1. Identify the population of interest and corresponding parameter
2. Obtain a suitable sample and an initial point estimate
3. Check validity assumptions to apply CLT
4. Calculate margin of error
5. Obtain lower and upper bounds
6. Interpret your $100(1-\alpha)\%$ interval estimate in the context of the problem

We will use the `pawnee` dataset in our class website to illustrate this procedure.

## Dataset Description

A brief description of the dataset

```{r}
pawnee <- read.csv(file="/Users/toledo60/Desktop/Projects/personal-website/data/pawnee.csv")
```


> The Sweetums candy factory in Pawnee, Indiana, is under investigation for violating EPA regulations. Factory workers have improperly disposed of arsenic and sulfur waste from the candy-making process, and the contamination has reached the local water supply! We have data for arsenic and sulfur levels from the water in all houses within a 2-mile radius of the factory


Some important variables include:

* `Arsenic`: arsenic levels for each home in ppm
* `Sulfur`: sulfur levels for each home in ppm
* `New_hlth_issue`: Indicates `Y` is someone living at the home has experienced a major health issue after the date of contamination, `N` if no new health complications

```{r}
head(pawnee)
```


We will construct a confidence interval for the population proportion of of households experiencing a major health issue caused by the EPA violations of the Sweetums candy factory in Pawnee, Indiana.


## Constructing a confidence interval


**Identify the population of interest and corresponding parameter**

Our population of interest is all houses within a 2-mile radius of Sweetums candy factory in Pawnee, Indiana.
The parameter is the proportion of houses within a 2-mile radius of Sweetums candy factory in Pawnee, Indiana which experienced a major health issue after the date of contamination

**Obtain a suitable sample and an initial point estimate**

Our sample will consist of 30 randomly chosen households from the `pawnee` dataset

```{r}
set.seed(1337)
rows_to_sample <- sample(1:nrow(pawnee), size = 30)
pawnee_sample <- pawnee[rows_to_sample,]
n <- nrow(pawnee_sample)
```

The initial point estimate, i.e the sample proportion of houses within a 2-mile radius of Sweetums candy factory in Pawnee, Indiana which experienced a major health issue after the date of contamination 

```{r}
p_hat <- mean(pawnee_sample$New_hlth_issue=="Y")
```

```{r}
p_hat
```


**Check validity assumptions to apply CLT**

```{r}
p_hat * n
(1-p_hat) * n
```

There were at least 10 failures but not at least 10 successes. However, we have a sufficiently large enough sample size $n \geq 30$ to apply the CLT. Therefore, the sampling distribution of the sample proportion is approximately normal

$$
\hat{p} \sim N \left(p, \sqrt{\frac{p(1-p)}{n}} \right)
$$

**Calculate margin of error**

The margin of error is defined as 

$$
ME = z_{1-\alpha/2} \times SE(\hat{p})
$$
We are interested in a 95\% confidence interval for the population proportion 


```{r}
alpha <- 0.05
critical_value <- qnorm(1-alpha/2)
SE <- sqrt(p_hat * (1-p_hat)/n)
```

Putting all of this together, we can compute the margin error 

```{r}
margin_error <- critical_value * SE
margin_error
```


**Obtain lower and upper bounds**

The lower bound is *LB = point estimate - margin of error*

```{r}
lb = p_hat - margin_error
```

The upper bound is *UB = point estimate + margin of error*

```{r}
ub = p_hat + margin_error
```


Therefore, our 95\% CI is 
```{r}
c(lb,ub)
```


**Interpret your $100(1-\alpha)\%$ interval estimate in the context of the problem**

We are 95\% confident the true population proportion of households within a 2-mile radius of Sweetums candy factory in Pawnee, Indiana which experienced a major health issue after the date of contamination is between 5.69\% and 34.31\%


Taking a look at the true population proportion 

```{r}
p <- mean(pawnee$New_hlth_issue == 'Y')
p
```

we see that our constructed confidence interval does contain the true population proportion

```{r}
p >= lb && p <= ub
```




